% !TEX root = ./multilinear.tex
\section{Introduction}

Many problems in graph mining and social network analysis can be reduced to questions
about different kinds of subgraphs. Two important classes of such problems, which 
are the focus of our paper, are:
(1) \emph{Detecting and counting subgraphs}, such as paths and trees, of a given size $k$---these are used
for characterizing different kinds of networks, especially in biological models
\cite{alon2008biomolecular, huffner2008algorithm}.
(2) \emph{Anomaly detection in network data using graph scan statistics}, which involves finding
connected subgraphs of size $k$, optimizing different kinds of anomaly score functions \cite{Speakman-14,leiserson2015pan,
hansen2016finding, \iftoggle{refsfull}{neil2013scan,} chen2014non}---this
arises in a number of applications, such as social network analysis, 
epidemiology, finance, and bio-surveillance \cite{akoglu2014graph}.

Both problems are computationally very challenging. For instance, exact detection of
paths is NP-hard and the corresponding counting problem is \#P-hard. Development
of parallel algorithms for these problems is an active area of research, and
many parallel algorithms exist for counting simple local subgraphs, such as triangles \cite{arif:cikm13, schmidt2009scalable, aparicio:ispa14, du:mcd09}. 
Finding trees is much harder, and a number of heuristics have been developed. One
of the few techniques that gives rigorous approximation guarantees is
\emph{color coding} \cite{alon2008biomolecular, alon1995color, huffner2008algorithm}.
Parallel adaptations have been developed using MapReduce \cite{zhao2012sahad} and
MPI \cite{slota:icpp13, slota:ipdps14}. The MPI-based 
FASCIA algorithm \cite{slota:icpp13, slota:ipdps14} is the state-of-the-art in terms of 
counting trees in massive networks, scaling to finding trees with up to 12 vertices in networks with
one billion edges. However, it seems very challenging to scale the color coding
method to larger subgraphs, even on small networks. The main reason is that the
time and space complexity of the color coding technique 
both scale as $2^k$, where $k$ denotes the subgraph size.
In this paper, we take the first steps towards beating this bound,
which has remained a significant open problem since \cite{slota:ipdps14}.
Our approach involves parallelization of a powerful algebraic technique for
detecting multilinear terms in a multivariate polynomial, developed by
Koutis \cite{koutis:icalp08} and Williams \cite{williams2009finding}.

Optimizing network scan statistics leads to challenging optimization problems as well.
Color coding has also been used to develop the first method with rigorous approximation
guarantees \cite{cadena:sdm17}; however, this has not been parallelized because of
its high memory overhead. In \cite{cadena:bigdata17}, we have developed a
parallel adaptation of the multilinear detection technique using both GraphX and Giraph. However,
none of these scaled beyond networks with 40 million edges.

In this paper, we develop a distributed algorithm for multilinear detection,
which immediately leads to highly scalable algorithms for both paths and trees,
and network scan statistics.
Our contributions are:

\noindent
\textbf{1. \ouralgo{}: Distributed algorithm for multilinear detection
and applications to subgraph analysis.}
We develop \ouralgo{}, a randomized distributed MPI based algorithm for finding
paths and trees through detection of
multilinear terms with $k$ variables of the form $x_{i_1}x_{i_2}\ldots x_{i_k}$ (i.e., a term in
which all variables have exponent 1) in a multivariate polynomial
$P(x_1,\ldots,x_n)$. The sequential algorithm uses a matrix representation
of a group algebra \cite{koutis:icalp08, williams2009finding}, and its structure lends itself to a natural
parallelization.  We give rigorous bounds on the performance in terms of the time and
space complexity, which scale as
$O(2^k)$ and $O(k)$, respectively, compared with $O(2^ke^k)$ and $O(2^k)$ for
color coding, respectively \cite{zhao2012sahad, slota:icpp13, slota:ipdps14}.
For random graphs, we show a rigorous scaling with $N$, the number of processors for $N\leq 2^k$.

\noindent
\textbf{2. Cache optimization and weak scalability}. %%Parallelization of the \textsc{$k$-MLD} problem {\color{red} TODO}
Our algorithm partitions the graph $G$ into $N_1$ parts. The computation involves $2^k$
iterations, and $N_1$ processors together perform one iteration---this allows us to schedule
$N/N_1$ such computations to occur in parallel. The total compute time exhibits good weak scaling.
Additionally, our data structures for supporting Galois field operations
during the iterations support a temporal cache locality, which actually leads to a reduction
in the compute time as $N_1$ increases. On the other hand, the communication cost increases
with $N_1$, leading to an optimal value of $N_1$ for the best performance.


\noindent
\textbf{3. Experimental results}. We evaluate our results on a number of real and synthetic networks with up to 250 million edges. The reduced memory footprint allows us to scale to finding subgraphs of size 18, which has not been done before.  Our algorithms for
both problems show reasonable scaling up to 512 processors, supporting our theoretical analysis. 
The running time grows linearly with the network size and as $2^k$ for the subgraph size $k$.

\noindent
\textbf{4. Comparison with prior methods.}
Our algorithm for finding paths is two orders of magnitude faster than FASCIA, the state of the art method based on color coding \cite{slota:icpp13, slota:ipdps14}.
Our algorithm for scan statistics improves on a Giraph-based implementation \cite{cadena:bigdata17} by over one order of magnitude, and it scales to significantly larger networks.

One additional advantage is that our parallel algorithm based on multilinear detection is conceptually much simpler than color coding based algorithms. It also requires far less bookkeeping than color coding. As we discuss in Section \ref{sec:proposed}, the obvious ways to try to parallelize multilinear detection do not scale well; instead, we find that a careful interplay between the degree of partitioning as well as batching a set of iterations yields better results.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%Many problems in graph mining and social network analysis can be reduced to questions
%%about different kinds of subgraphs; two important classes of such problems, which 
%%are the focus of our paper, are:
%%(1) \emph{Detecting subgraphs}, such as paths and trees, of a given size $k$---these are used
%%for characterizing different kinds of networks, especially in biological models
%%\cite{alon2008biomolecular, huffner2008algorithm}.
%%(2) \emph{Anomaly detection in networked data using graph scan statistics}, which involves finding
%%connected subgraphs of size $k$, optimizing different kinds of objectives \cite{Speakman-14,leiserson2015pan,
%%hansen2016finding, \iftoggle{refsfull}{neil2013scan,} chen2014non}.
%%This problem arises in a number of applications, such as social network analysis, 
%%epidemiology, finance, and bio-surveillance. We refer to the survey by 
%%Akoglu et al. \cite{akoglu2014graph} for a detailed discussion on different approaches 
%%to graph anomaly detection.
%%
%%The problems of subgraph detection and scan statistics are NP-hard, in general, making them 
%%computationally very challenging. Various sequential heuristics have been proposed for
%%both these problem classes, e.g.,
%%\cite{alon2008biomolecular, huffner2008algorithm, Speakman-14,leiserson2015pan,
%%hansen2016finding, \iftoggle{refsfull}{neil2013scan,} chen2014non, cheng:kdd12}. However, these
%%do not scale to large instances, motivating the need for parallel implementations.
%%There has been a fair amount of work on parallel algorithms for various kinds of subgraph
%%problems, e.g., \cite{zhao2012sahad, slota:icpp13, slota:ipdps14, arif:cikm13, 
%%schmidt2009scalable, zhao2016parallel, aparicio:ispa14, du:mcd09}, but very little for
%%anomaly detection using graph scan statistics \cite{zhao2016parallel}.
%%
%%
%%Most of these parallel algorithms involve heuristics based on partitioning, pruning, and 
%%enumeration for speeding up the computations to fairly large graphs---e.g., for 
%%maximal cliques \cite{schmidt2009scalable, zhao2016parallel, aparicio:ispa14, cheng:kdd12, du:mcd09}.
%%One of the few rigorous approaches for subgraph counting that has been used for
%%developing parallel algorithms is based on the ``color coding'' technique
%%\cite{alon2008biomolecular, alon1995color}, which gives a rigorous \emph{fixed parameter tractable} algorithm\footnote{This means the running time is $O(f(k)n^c)$, where
%%$n$ is the number of nodes, $c$ is a constant, and $f(\cdot)$ is a function of some other
%%parameter---in this case, it is the size of the subgraph \cite{downey}.}
%%for finding and counting subgraphs. Color coding is a dynamic programming algorithm,
%%and multiple approaches have been developed for parallelizing it 
%%\cite{zhao2012sahad, slota:icpp13, slota:ipdps14}; the FASCIA algorithm of \cite{slota:icpp13, slota:ipdps14}
%%which is implemented in MPI is the state-of-the-art, and can handle trees
%%of size up to 12 in graphs with millions of nodes. 
%%%%Color coding has also been used to develop a
%%%%sequential algorithm for graph scan statistics \cite{cadena:sdm17}, but no parallel
%%%%adaptations exist so far.
%%A big challenge of the color coding technique is the memory overhead, since
%%it scales as $2^k$, where $k$ denotes the subgraph size, which limits further
%%scaling to larger graphs or subgraphs.  
%%
%%In this paper, we develop a novel approach for designing highly scalable parallel
%%algorithms for both the problem classes mentioned above, by adapting
%%powerful algebraic techniques for
%%detecting multilinear terms in a multivariate polynomial developed by 
%%Koutis \cite{koutis:icalp08} and Williams \cite{williams2009finding}.
%%Our contributions are:
%%
%%\noindent
%%\textbf{1. Efficient algorithms for parallel multilinear detection
%%and applications to subgraph analysis.}
%%We develop a distributed MPI based algorithm for detecting 
%%multilinear terms with $k$ variables of the form $x_{i_1}x_{i_2}\ldots x_{i_k}$ (i.e., a term in
%%which all variables have exponent 1) in a multivariate polynomial
%%$P(x_1,\ldots,x_n)$---this is referred to as the $k$-Multilinear Detection (\textsc{$k$-MLD}) problem
%%\cite{koutis:icalp08, williams2009finding}. The problem of finding paths and trees can
%%be reduced to the \textsc{$k$-MLD} problem \cite{koutis:icalp08}. We also show that the
%%network scan statistics problem can also be reduced to the \textsc{$k$-MLD} problem.
%%Consequently, our parallel multilinear detection algorithm leads to parallel algorithms
%%for both these problems. Our methods provably
%%have better space and running time complexity compared to the color coding technique.
%%The space complexity is $O(kn)$ compared to $O(2^kn)$, whereas the running time complexity is proportional to $O(2^k)$, compared to $O(2^ke^k)$
%%\cite{zhao2012sahad, slota:icpp13, slota:ipdps14}.
%%
%%\noindent
%%\textbf{2. Experimental results}. We evaluate our results on a number of real and synthetic networks 
%%with up to 250 million edges and subgraph sizes up to $k=18$. The reduced memory footprint
%%allows us to scale to paths of size 18, which has not been done before.  Our algorithms for
%%both problems show reasonable scaling up to 512 processors. The running time grows linearly
%%with the network size and as $2^k$ for the subgraph size $k$.
%%
%%\noindent
%%\textbf{3. Comparison with prior methods.}
%%Our algorithm for finding paths gives over two orders of magnitude improvement in time
%%compared to FASCIA, the state of the art method based on color coding \cite{slota:icpp13, slota:ipdps14}.
%%There is only one parallel algorithm for network scan statistics, which is based on pruning heuristics,
%%and our method gives several order of magnitude improvement over it.
%%
%%\noindent
%%\textbf{4. Comparison with Giraph.}
%%We compare our methods with recent parallel graph programming models, Giraph and Spark.
%%Our results show that neither of these models scale well for these problems; further,
%%Spark has worse performance than Giraph.


%we study a novel approach for a broad class of subgraph analysis problems
%in massive networks based on powerful algebraic techniques. This involves formalizing these
%problems in terms of detecting multilinear terms in a multivariate polynomial, based on 
%very powerful techniques developed by Koutis \cite{koutis:icalp08} and Williams \cite{williams2009finding}.
%However, only sequential algorithms have been developed so far \cite{bjorklund:esa14}, which do not
%scale beyond networks with $\sim 10^3-10^4$ nodes.
%We develop a novel approach for parallel multilinear detection, which allows us to
%solve a broad class of subgraph analysis and anomaly detection problems with the same tools.
%
%Both multilinear detection and subgraph counting using color coding are examples of
%\emph{fixed parameter tractable} algorithms, which run in time $O(f(k)n^c)$, where
%$n$ is the number of nodes, $c$ is a constant, and $f(\cdot)$ is a function of some other
%parameter---in this case, it is the size of the subgraph. Most of these problems are
%NP-complete, in general, and such algorithms are one way to handle the hardness; this is
%referred to as \emph{parameterized complexity} (see \cite{downey} for an excellent discussion
%of such methods for many fundamental NP-hard problems).
%
%In this paper, we study parallel algorithms for multilinear detection and their applications
%to different subgraph analysis problems, using multiple models of parallel graph computation.
%Our contributions are:

